Modularly Composing Typed Language Fragments

### 2014-11-02 att-pldi15/att-pldi15.tex
Edits to related work

The fifth premise of (syn-targ) extracts the \emph{targeted opcon definition} of $\opname{op}$ from the opcon structure, $\omega$. Like the intro opcon definition, this is a static function that generates a translational internal term on the basis of the target tycon's type index, the term index and an argument interface list. Targeted opcon definitions additionally synthesize a type. The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that no tycon defines an opcon twice and that the opcon definitions are well-kinded. 
For example, $\omega_\text{rstr/targops}$ defines:
\[\small
\begin{array}{l}
\keyw{syn}~\opname{conc}=\slam{\krx}{\svar{tyidx}}{
    \slam{\kunit}{\svar{tmidx}}{
        \slam{\kargs}{\svar{args}}{\\
            \quad \keyw{letpair}~(\svar{arg1}, \svar{arg2}) = \svar{arity2}~\svar{args}\\
            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}\\
            \quad \keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\
            \quad \stycase{\tcvar{rstr}}{\svar{ty2}}{\svar{tyidx2}}{\\
                \quad\quad (\sty{\tcvar{rstr}}{\svar{rxconcat}~\svar{tyidx}~\svar{tyidx2}}, \sqitm{sconcat~\quq{\svar{tr1}}~\quq{\svar{tr2}}})\\
                \quad
            }{\sraise{\kty\times\kitm}}
        }
    }
}
\end{array}
\]
The helper function $\svar{arity2}$ checks that two arguments, including the target argument, were provided. We then request synthesis of both arguments. We can ignore the type synthesized by the first because by definition it is a regular string type with type index $\svar{tyidx}$. We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and raising an error if not. We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type index, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.

@@

The reason why the rule (syn-targ-other) is never a problem in proving a tycon invariant turns out to be the same reason extending the tycon context conserves all tycon invariants. A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g. $\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g. $\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check the translation against $\tau_\text{abs}=\alpha$, which would fail. %Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
 We can state this more generally:

@@

%\paragraph{Refinement Types}
%When T, the solution is less clear. 
When new static distinctions are needed within an existing type, but new operators are not necessary and cost is acceptable, one solution is to develop a system of \emph{refinement types} for the language \cite{Freeman91}. For example, a refinement of  integers might distinguish negative integers. Proposals for ``pluggable type systems'' describe such type refinement systems \cite{Brac04a,Andreae:2006:FIP:1167473.1167479}. Refinements of abstract types can be used to hold representations abstract. %Because the dynamic semantics are fixed,  composition is safe. Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.

@@

Proof assistants and logical frameworks can be used to mechanize dialect specification and metatheory, but they similarly require a complete specification. Chlipala has suggested using  proof automation that covers a variety of language features to decrease the burden when a  dialect is formed, but this is a fundamentally heuristic approach \cite{Chlipala10}\todo{wrong citation}. Our approach provides a conservativity theorem. No proof terms are needed to guarantee modular reasoning:   mistakes can only weaken the invariants at a single tycon. %This facilitates exploratory language design and suggests educational applications.  
A promising avenue for future work beyond the scope of this paper is to introduce optional proof mechanization into the SL (by making it dependently kinded), and to mechanize the metatheory of @$\lambda$ itself.%Every combination of fragments is a new dialect, and must be reasoned about monolithically. %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so . 
% Let us briefly review difficulties that arise.

@@

Type abstraction is, encouragingly, also the technique underlying modular reasoning in ML-like languages. Indeed, proofs of tycon invariants rely on the same abstraction theorem. Our work has some technical similarities to work on syntactic type abstraction \cite{journals/toplas/GrossmanMZ00} and on translating modules to System F \cite{conf/tldi/RossbergRD10}.  Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) can be \emph{computed} based on a type index. Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon. Their static and dynamic semantics can be complex because a static function computes the type and translation. In contrast, modules can only specify a fixed number of operations, and each must have function type. Note that although we did not specify quantification over external types here for simplicity, we conjecture that it is complementary (and thus @$\lambda$ could serve as the core of a language with a ML-style module system). Another avenue for future work is tycon functors, which abstract over tycons with the same tycon signature. % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions. %Both examples we used here would be quite difficult to fully embed using only modules.

>>>>

The fifth premise of (syn-targ) extracts the \emph{targeted opcon definition} of $\opname{op}$ from the opcon structure, $\omega$. Like the intro opcon definition, this is a static function that generates a translational internal term on the basis of the target tycon's type index, the term index and an argument interface list. Targeted opcon definitions additionally synthesize a type. The rule (ocstruct-targ) in Figure \ref{ocstruct} ensures that no tycon defines an opcon twice and that the opcon definitions are well-kinded. 
For example, $\omega_\text{rstr/targops}$ defines:
\[\small
\begin{array}{l}
\keyw{syn}~\opname{conc}=\slam{\krx}{\svar{tyidx}}{
    \slam{\kunit}{\svar{tmidx}}{
        \slam{\kargs}{\svar{args}}{\\
            \quad \keyw{letpair}~(\svar{arg1}, \svar{arg2}) = \svar{arity2}~\svar{args}~\keyw{in}\\
            \quad \keyw{letpair}~(\_, \svar{tr1}) = \svar{syn}~\svar{arg1}~\keyw{in}~ \keyw{letpair}~(\svar{ty2}, \svar{tr2}) = \svar{syn}~\svar{arg2}\\
            \quad \stycase{\tcvar{rstr}}{\svar{ty2}}{\svar{tyidx2}}{\\
                \quad\quad (\sty{\tcvar{rstr}}{\svar{rxconcat}~\svar{tyidx}~\svar{tyidx2}}, \sqitm{sconcat~\quq{\svar{tr1}}~\quq{\svar{tr2}}})\\
                \quad
            }{\sraise{\kty\times\kitm}}
        }
    }
}
\end{array}
\]
The helper function $\svar{arity2}$ checks that two arguments, including the target argument, were provided. We then request synthesis of both arguments. We can ignore the type synthesized by the first because by definition it is a regular string type with type index $\svar{tyidx}$. We case analyze the second against $\tcvar{rstr}$, extracting its index regular expression if so and raising an error if not. We then synthesize the resulting regular string type, using the helper function $\svar{rxconcat} :: \krx \rightarrow \krx \rightarrow \krx$ which generates the synthesized type index, and finally the translation, using an internal helper function $sconcat : \keyw{str} \rightarrow \keyw{str} \rightarrow \keyw{str}$, the translational term for which we assume has been substituted in directly.

@@

The reason why the rule (syn-targ-other) is never a problem in proving a tycon invariant, i.e. \emph{translation independence} of tycons, turns out to be the same reason extending the tycon context conserves all tycon invariants. A newly introduced tycon defining a targeted operator that synthesizes a regular string type, e.g. $\stx{paper}$, and generating a translation that is not in the corresponding regular language, e.g. $\texttt{""}$, could be defined, but when used, the rule (syn-targ) would check the translation against $\tau_\text{abs}=\alpha$, which would fail. %Type abstraction is the basis for conservatively composing type system fragments, just like it is the basis for composing ML-style modules.
 We can state this more generally:

@@

%\paragraph{Refinement Types}
%When T, the solution is less clear. 
When new static distinctions are needed within an existing type, but new operators are not necessary and cost is acceptable, one solution is to develop a system of \emph{refinement types} for the language \cite{Freeman91}. For example, a refinement of  integers might distinguish negative integers. Proposals for ``pluggable type systems'' describe composing such type refinement systems \cite{Brac04a,Andreae:2006:FIP:1167473.1167479}. Refinements of abstract types can be used to hold representations abstract. %Because the dynamic semantics are fixed,  composition is safe. Although our regular string types could perhaps be approximated with refinements, this does not permit the advanced operations, optimizations and representation independence results we enable.

@@

Proof assistants and logical frameworks can be used to mechanize dialect specification and metatheory, but they similarly require a complete specification. Chlipala has suggested using  proof automation that covers a variety of language features to decrease the burden when forking a  dialect, but this is a fundamentally heuristic approach \cite{Chlipala10}\todo{wrong citation}. Our approach provides a conservativity theorem. No proof terms are needed to guarantee modular reasoning:   mistakes in a tycon definition can only weaken invariants at that tycon. %This facilitates exploratory language design and suggests educational applications.  
A promising avenue for future work is to introduce optional proof mechanization into the SL (by making it dependently kinded), and to mechanize the metatheory of @$\lambda$ itself.%Every combination of fragments is a new dialect, and must be reasoned about monolithically. %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so . 
% Let us briefly review difficulties that arise.

@@

Type abstraction is, encouragingly, also the technique underlying modular reasoning in ML-like languages\todo{citation}. Indeed, proofs of tycon invariants can rely on the same abstraction theorem. Our work has some technical similarities to work on syntactic type abstraction \cite{journals/toplas/GrossmanMZ00} and on translating modules to System F \cite{conf/tldi/RossbergRD10}.  Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) can be \emph{computed} based on a type index. Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon. Their static and dynamic semantics can be complex because a static function computes the type and translation. In contrast, modules can only specify a fixed number of operations, and each must have function type. Note that although we did not specify quantification over external types here for simplicity, we conjecture that it is complementary (and thus @$\lambda$ could serve as the core of a language with a ML-style module system). Another avenue for future work is tycon functors, which abstract over tycons with the same tycon signature. % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions. %Both examples we used here would be quite difficult to fully embed using only modules.
